Breadth First Searches Over Evolving Graphs

### 2016-01-27 paper.tex
Add comment about nilpotence and termination

For each static graph $G^{[t]} = (V^{[t]}, E^{[t]})$ that constitutes the
evolving graph $G_n$, define its corresponding
$\left\vert V^{[t]}\right\vert \times \left\vert V^{[t]}\right\vert$
adjacency matrix with elements
\[
A^{[t]}_{ij} =
\begin{cases}
 1 & \mbox{if $(i,j)\in E^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\]

@@

The off-diagonal blocks $M^{[t_i, t_j]}$ provide an explicit matrix representation
for the $\odot$ product in that $(M^{[t_i, t_j]})^T b = A^T \odot b$. The upper
triangular structure of $\bm M_n$ (and hence $\bm A_n$) reflects the causal
nature of temporal paths in that they cannot go backward in time.

@@

For the example of Figure~\ref{fig:eg_shortest_path}, we have
\begin{align*}
V & = \{(1,\!t_1), (2,\!t_1), (1,\!t_2), (3,\!t_2), (2,\!t_3), (3,\!t_3)\},\\
\cup_t\!E^{[t]} &= \{((1,\!t_1), (2,\!t_1)), ((1,\!t_2), (3,\!t_2)), ((2,\!t_3), (3,\!t_3))\},\\
E'  &= \{((1,\!t_1), (1,\!t_2)), ((2,\!t_2), (2,\!t_3)), ((3,\!t_2), (3,\!t_3))\}.
\end{align*}
In the order specified for $V$, the adjacency matrix of $G$ is then
\[
\bm A_3 = \begin{bmatrix}
0 & 1 & 1 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 \\
0 & 0 & 0 & 0 & 0 & 1 \\
0 & 0 & 0 & 0 & 0 & 0
\end{bmatrix}
\]

@@

We see that $\bm (A^T)^2$ correctly encodes the collection of products in
\eqref{eq:5}-\eqref{eq:6}. Furthermore, $\bm (A^T)^3 \bm b$ correctly counts the
two allowed temporal paths from $(1, t_1)$ to $(3, t_3)$, and that the
off-diagonal structure encoded in $E'$ and the $M^{[t,t']}$ blocks are critical
to obtaining the correct count.

@@

The blocked matrix--vector products introduced in the previous section allows us
to write down an elegant algebraic formulation of BFS on evolving graphs, as
presented in Algorithm~\ref{alg:bfsla}. Note that $\bm A^T$ need never be formed
explicitly, as only its matrix--vector product is required in the algorithm.

\begin{algorithm}[h]
 \SetKwFor{For}{for}{}{end}
 \SetKwFor{While}{while}{}{end}
 \SetKwFor{If}{if}{}{end}
 \SetKwFunction{nei}{forward-neighbors}
 \SetKwFunction{enqueue}{enqueue}
 \SetKwFunction{nz}{nonzeros}
 \SetKwFunction{intersect}{intersect}
 \SetKwFunction{map}{map}
 \SetKwProg{Fn}{function}{}{end}
 \Fn{ABFS($A_n, (v_1,t_1)$)}{
   Form $\bm{A}^T$ from $A_n$.\\
   $\bm{b}$ is a sparse vector of length $mn$. \\
   $\bm{b}_s = 1$ \\
   $k = 1$ \\
   $reached[(v_1,t_1)] = 0$ \\
   \While{\nz{$\bm{b}$} $\ne \emptyset$}{
       $\bm{b} = \bm{A}^T\bm{b}$ \\
       \For{$k \in$ \nz{$\bm{b}$}}{
         \If{\map{$k$} $\in reached$}{
           $\bm{b}_k = 0$ \\
         }
      }
      $activeNodes =$\map{$\bm{b}$}\\
       \For{$activeNode \in activeNodes$}{
           $reached[activeNode] = k$ \\
         }
       $k = k + 1$
       }
  \textbf{return} $reached$ \\
}
\caption{An algebraic formulation of BFS on evolving graphs.
Given $A_n$, the adjacency matrix representation of $G_n$ and $(v_1,t_1)$, a
node of $G_n$, returns $reached$ as defined in Algorithm~\ref{alg:bfs}.
The function \texttt{nonzeros(v)} returns the nonzero indices of the vector $v$,
and the function \texttt{map(b)} maps a block vector's index (or indices) to their corresponding
active node (or active nodes).}
\label{alg:bfsla}
\end{algorithm}

>>>>

For each static graph $G^{[t]} = (V^{[t]}, E^{[t]})$ that constitutes the
evolving graph $G_n$, define its corresponding
$\left\vert V^{[t]}\right\vert \times \left\vert V^{[t]}\right\vert$
adjacency matrix with elements
\begin{equation}
A^{[t]}_{ij} =
\begin{cases}
 1 & \mbox{if $(i,j)\in E^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\label{eqn:adjacency}
\end{equation}

@@

The off-diagonal blocks $M^{[t_i, t_j]}$ provide an explicit matrix representation
for the $\odot$ product in that $(M^{[t_i, t_j]})^T b = (A^{[t_i]})^T \odot b$.
The upper triangular structure of $\bm M_n$ (and hence $\bm A_n$) reflects the causal
nature of temporal paths in that they cannot go backward in time.

@@

%XXX Would be nice to have a figure for $G$.
For the example of Figure~\ref{fig:eg_shortest_path}, we have
\begin{align*}
V & = \{(1,\!t_1), (2,\!t_1), (1,\!t_2), (3,\!t_2), (2,\!t_3), (3,\!t_3)\},\\
\cup_t\!E^{[t]} &= \{((1,\!t_1), (2,\!t_1)), ((1,\!t_2), (3,\!t_2)), ((2,\!t_3), (3,\!t_3))\},\\
E'  &= \{((1,\!t_1), (1,\!t_2)), ((2,\!t_2), (2,\!t_3)), ((3,\!t_2), (3,\!t_3))\}.
\end{align*}
In the order specified for $V$, the adjacency matrix of $G$ is then
\[
\bm A_3 = \begin{bmatrix}
0 & 1 & 1 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 \\
0 & 0 & 0 & 0 & 0 & 1 \\
0 & 0 & 0 & 0 & 0 & 0
\end{bmatrix}
\]

@@

We see that $\bm (A_n^T)^2$ correctly encodes the collection of products in
\eqref{eq:5}-\eqref{eq:6}. Furthermore, $\bm (A_3^T)^3 \bm b$ correctly counts the
two allowed temporal paths from $(1, t_1)$ to $(3, t_3)$, and that the
off-diagonal structure encoded in $E'$ and the $M^{[t,t']}$ blocks are critical
to obtaining the correct count.

Finally, we prove a simple lemma that the block adjacency matrix $\bm A_n$ is
nilpotent whenever all the subgraphs $G^{[t]}$ of $G_n$ are acyclic.

\begin{lemma}[Acyclicity implies nilpotence]
Let $G_n = \langle G^{[t_i]} \rangle_{i=1}^n$ be an evolving directed graph and
let all the directed graphs $G^{[t]}$ be acyclic. Then $\bm A_n$ is nilpotent.
\label{thm:nilpotent}
\end{lemma}

\IEEEproof
Since each directed graph $G^{[t]}$ is acyclic, its corresponding adjacency
matrix $A^{[t]}$ is strictly upper triangular. As a result, $\bm A_n$ must be
upper triangular.

Furthermore, none of the graphs $G^{[t]}$ can have any self-edges,
i.e.\ edges of the form $(u, u)$, and so all diagonal entries of
$A^{[t]}$ must be zero.  Therefore all the diagonal entries of $\bm A_n$ by
construction must be zero also.

We have now proven that $\bm A_n$ is an upper triangular matrix whose diagonal
entries are all zero. Therefore, $\bm A_n$ is nilpotent.
\endproof

Lemma~\ref{thm:nilpotent} also holds for acyclic undirected graphs so long as
the corresponding adjacency matrix representation is encoded in an asymmetric
fashion akin to \eqref{eqn:adjacency}.

@@

The blocked matrix--vector products introduced in the previous section allows us
to write down an elegant algebraic formulation of BFS on evolving graphs, as
presented in Algorithm~\ref{alg:bfsla}. Note that $\bm A_n^T$ need never be formed
explicitly, as only its matrix--vector product is required in the algorithm.

\begin{algorithm}[h]
 \SetKwFor{For}{for}{}{end}
 \SetKwFor{While}{while}{}{end}
 \SetKwFor{If}{if}{}{end}
 \SetKwFunction{nei}{forward-neighbors}
 \SetKwFunction{enqueue}{enqueue}
 \SetKwFunction{nz}{nonzeros}
 \SetKwFunction{intersect}{intersect}
 \SetKwFunction{map}{map}
 \SetKwProg{Fn}{function}{}{end}
 \Fn{ABFS($A_n, (v_1,t_1)$)}{
   Form $\bm A^T_n$ from $A_n$.\\
   $\bm{b}$ is a sparse vector of length $mn$. \\
   $\bm{b}_s = 1$ \\
   $k = 1$ \\
   $reached[(v_1,t_1)] = 0$ \\
   \While{\nz{$\bm{b}$} $\ne \emptyset$}{
       $\bm{b} = \bm{A}^T_n\bm{b}$ \\
       \For{$k \in$ \nz{$\bm{b}$}}{
         \If{\map{$k$} $\in reached$}{
           $\bm{b}_k = 0$ \\
         }
      }
      $activeNodes =$\map{$\bm{b}$}\\
       \For{$activeNode \in activeNodes$}{
           $reached[activeNode] = k$ \\
         }
       $k = k + 1$
       }
  \textbf{return} $reached$ \\
}
\caption{An algebraic formulation of BFS on evolving graphs.
Given $A_n$, the adjacency matrix representation of $G_n$ and $(v_1,t_1)$, a
node of $G_n$, returns $reached$ as defined in Algorithm~\ref{alg:bfs}.
The function \texttt{nonzeros(v)} returns the nonzero indices of the vector $v$,
and the function \texttt{map(b)} maps a block vector's index (or indices) to their corresponding
active nodes.}
\label{alg:bfsla}
\end{algorithm}

\begin{thm}
Algorithm~\ref{alg:bfsla} terminates.
\end{thm}

\IEEEproof
Recall from Lemma~\ref{thm:nilpotent} that $\bm A_n$ is nilpotent, i.e.\
there exists some positive integer $k$ for which $\bm A_n^k = 0$.
Hence, after iteration $k$, $\bm b$ is assigned the value $\bm A_n^k \bm b = 0$.
Therefore, Algorithm~\ref{alg:bfsla} must terminate after iteration $k$.
\endproof
